---
title: "Report on Diabetes Behavioral Factors Analysis"
author: "Anupama Bishwokarma"
format: html
date: today
editor: visual
code-fold: true
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(
  echo = TRUE,       # Show code
  warning = FALSE,   # Hide warnings
  message = FALSE    # Hide messages
)
```

## Project overview

Diabetes behavioral factors analysis project analyzes the responses collected among the adults in the United States using Behavioral Risk Factor Surveillance System (BRFSS) in 2023. The project is using on R programming language.

## Project objectives

1.  To explore the socio-demographic, biological and behavioral characteristics of the respondents
2.  To examine the association of behavioral factors with the diabetes status

## BRFSS dataset overview

BRFSS collects uniform state-specific data on health risk behaviors, chronic diseases and conditions, access to health care, and use of preventive health services related to the leading causes of death and disability in the United States. Data Source: <https://www.cdc.gov/brfss/annual_data/annual_2023.html> under the file name \[2023 BRFSS Data (SAS Transport Format)\].

## Load library
```{r}
# Load all required packages
library("Hmisc")    # to assign labels to variables
library("ggplot2")  # for visualization of graphs
library("usmap")    # to visualize the USA map
library("jtools")   # to produce summary of regression model
```

## Load dataset

```{r}
# Loading .XPT format dataset
# Issue: Large file size (> 1 GB), did not push to Github
# whole_brfss <- read_xpt("01_Datasets/whole_BRFSS_2023.XPT") # package "haven" used to read .XPT file

# Solution: Saved to .rds format compressed file (45.8 MB)
# saveRDS(whole_brfss, 
#        file = "./01_Datasets/whole_BRFSS_2023.rds", 
#        compress = TRUE)

# Load .rds format file
whole_brfss <- readRDS("01_Datasets/whole_BRFSS_2023.rds")
```

```{r}
#| echo: false
#| output: false
# Quick snapshot of the whole dataset
# knitr::kable(head(whole_brfss))
```

## Data preparation: subsetting, labelling, recoding, standardizing

The loaded dataset was explored for relevant variables for the project. Additionally, I referred to questionnaire <https://www.cdc.gov/brfss/questionnaires/pdf-ques/2023-BRFSS-Questionnaire-508.pdf> used for the survey to understand the topic and questions. The whole dataset included 350 variables with 400,000+ rows.

The dataset was filtered for the relevant 12 variables with same number of observations. These variables included: socio-demographic, biological, behavioral and diabetes related variables. The outcome variable is of "Diabetes Status".

```{r}
#| echo: false
#| output: false
# Create dataframe with objectives relevant variables 
short_brfss <- whole_brfss[c("SEQNO", 
                             "_STATE", 
                             "_AGE80", 
                             "SEXVAR", 
                             "_INCOMG1", 
                             "_BMI5", 
                             "EXERANY2",
                             "_RFSMOK3", 
                             "USENOW3", 
                             "DRNKANY6", 
                             "DIABETE4",
                             "DIABAGE4")]
```

Then, the selected variables were renamed for easy readability.

```{r}
# Renaming the variables: 
names(short_brfss) <- c("id_no", 
                        "state_name",        # Name of the State
                        "age",               # Age of the respondent
                        "sex",               # Sex of the respondent
                        "income_level",      # Annual Household income from all sources
                        "BMI",               # Body Mass Index
                        "physical_activity", # Any physical activity in past month?
                        "smoking",           # Current smoking status
                        "tobacco_use",       # Current tobacco use
                        "alc_drnk_30days",   # At least one drink of alcohol in the past 30 days
                        "diabetes_status",   # Diabetes status
                        "diab_first_known")  # Age when first told had diabetes
```

Next, dataset was checked for the type of variables. The dataset included of numeric and character data types. Categorical variables were integer-encoded.

```{r}
#| echo: false
#| output: false
# Check for data types
str(short_brfss)
```

Following this, categorical variables were converted to labeled factors. The levels "Yes" and "No" of "smoking" were reordered for uniformity. Additionally, levels of certain variables (e.g, "tobacco_use", "diabetes_status" ) were re-classified to two-levels.

```{r}
# assign labels to variable: state_name
state_labels <- c("1" = "Alabama", "2" = "Alaska", "4" = "Arizona", 
                  "5" = "Arkansas", "6" = "California", "8" = "Colorado", 
                  "9" = "Connecticut", "10" = "Delaware", 
                  "11" = "District of Columbia", "12" = "Florida", 
                  "13" = "Georgia", "15" = "Hawaii", "16" = "Idaho", 
                  "17" = "Illinois", "18" = "Indiana", "19" = "Iowa", 
                  "20" = "Kansas", "22" = "Louisiana", "23" = "Maine", 
                  "24" = "Maryland", "25" = "Massachusetts", "26" = "Michigan",
                  "27" = "Minnesota", "28" = "Mississippi", "29" = "Missouri", 
                  "30" = "Montana", "31" = "Nebraska", "32" = "Nevada", 
                  "33" = "New Hampshire", "34" = "New Jersey", 
                  "35" = "New Mexico", "36" = "New York", "37" = "North Carolina",
                  "38" = "North Dakota", "39" = "Ohio", "40" = "Oklahoma", 
                  "41" = "Oregon", "44" = "Rhode Island", 
                  "45" = "South Carolina", "46" = "South Dakota", 
                  "47" = "Tennessee", "48" = "Texas", "49" = "Utah", 
                  "50" = "Vermont", "51" = "Virginia", "53" = "Washington", 
                  "54" = "West Virginia", "55" = "Wisconsin", "56" = "Wyoming", 
                  "66" = "Guam", "72" = "Puerto Rico", "78" = "Virgin Islands"
                  )


short_brfss$state_name <- factor(short_brfss$state_name,
                                 levels = as.numeric(names(state_labels)),
                                 labels = state_labels
                                 )


# assign labels to variable: sex
short_brfss$sex <- factor(short_brfss$sex, 
                              levels = c(1, 2),
                              labels = c("Male", "Female"))


# assign labels to variable: income_level
income_labs <- c("1" = "Less than $15,000",
                         "2" = "$15,000 to < $25,000",
                         "3" = "$25,000 to < $35,000",
                         "4" = "$35,000 to < $50,000",
                         "5" = "$50,000 to < $100,000",
                         "6" = "$100,000 to < $200,000",
                         "7" = "$200,000 or more",
                         "9" = "Don't know/Not sure/Missing")


short_brfss$income_level<- factor(short_brfss$income_level,
                                  levels = as.numeric(names(income_labs)),
                                  labels = income_labs)


# assign labels to variable: physical_activity
short_brfss$physical_activity <- factor(short_brfss$physical_activity,
                                            levels = c(1, 2, 7, 9),
                                            labels = c("Yes",
                                                       "No",
                                                       "Don't know/Not Sure",
                                                       "Refused"))


# assign labels to variable: smoking
short_brfss$smoking <- factor(short_brfss$smoking,
                                      levels = c(1, 2, 9),
                                      labels = c("No",
                                                 "Yes",
                                                 "Don't know/Refused/Missing"))


#Reorder the levels of smoking
short_brfss$smoking <- factor(short_brfss$smoking, levels = c("Yes", "No"))



# assign labels to variable: tobacco_use
short_brfss$tobacco_use <- factor(short_brfss$tobacco_use,
                              levels = c(1, 2, 3, 7, 9),
                              labels = c("Every day",
                                         "Some days",
                                         "Not at all",
                                         "Don't know/Not Sure",
                                         "Refused"))



# Modification to classify current tobacco users into two levels
short_brfss$tobacco_use <- ifelse(short_brfss$tobacco_use %in% 
                                    c("Every day", "Some days"), "Yes",
                           ifelse(short_brfss$tobacco_use == "Not at all", "No", NA))



# assign labels to variable: alc_drnk_30days
short_brfss$alc_drnk_30days <- factor(short_brfss$alc_drnk_30days,
                                      levels = c(1, 2, 7, 9),
                                      labels = c("Yes",
                                                 "No",
                                                 "Don't know/Not Sure",
                                                 "Refused"))


# assign labels to variable: diabetes_status
short_brfss$diabetes_status <- factor(short_brfss$diabetes_status,
                                      levels = c(1, 2, 3, 4, 7, 9),
                                      labels = c("Yes",
                                                 "Gestational diabetes",
                                                 "No",
                                                 "Pre-diabetes",
                                                 "Don't know/Not Sure",
                                                 "Refused"))


```

Example: In variable "diabetes_status", "Gestational diabetes" was re-classified to "Yes" and "Pre-diabetes" to "No". This would be crucial for later logistics regression as in R it needs outcome variable "Y" with a factor with two levels.

```{r}
# Re-classifying into two-factor levels: diabetes_status
short_brfss$diabetes_status[short_brfss$diabetes_status == 
                              "Gestational diabetes"] <- "Yes"
short_brfss$diabetes_status[short_brfss$diabetes_status == 
                              "Pre-diabetes"] <- "No"
```

Following code adds 2 decimal places for BMI variable.

```{r}
# Adding 2 decimal places for numerical variable: BMI
short_brfss$BMI <- short_brfss$BMI/100
```

Categorical variables had factor levels like missing, don't know/not sure, refused. This was converted to NA for easier missing value calculation.

```{r}
# Function prep to replace with NA
to_be_replaced <- list(
  income_level = c("Don't know/Not sure/Missing"),
  physical_activity = c("Don't know/Not Sure","Refused"),
  smoking         = c("Don't know/Refused/Missing"),
  alc_drnk_30days = c("Don't know/Not Sure","Refused"),
  diabetes_status = c("Don't know/Not Sure", "Refused")
  )

replace_with_na <- function(df, to_be_replaced) {
  for (v in names(to_be_replaced)) 
    {
    df[[v]][df[[v]] %in% to_be_replaced[[v]]] <- NA
    }
  return(df)
  }


short_brfss <- replace_with_na(short_brfss, to_be_replaced)
```

```{r}
#| echo: false
#| output: false
# Deleting empty factor levels from the variables for keeping only filled levels
short_brfss$income_level <- factor(short_brfss$income_level, 
                                   levels = c( "Less than $15,000",
                                               "$15,000 to < $25,000",
                                               "$25,000 to < $35,000",
                                               "$35,000 to < $50,000",
                                               "$50,000 to < $100,000",
                                               "$100,000 to < $200,000",
                                               "$200,000 or more"))


# Reusing Yes, No levels
reuse_yn <- c("Yes", "No")


short_brfss$physical_activity <- factor(short_brfss$physical_activity, 
                                        levels = reuse_yn)


short_brfss$smoking <- factor(short_brfss$smoking, 
                              levels = reuse_yn)


short_brfss$tobacco_use <- factor(short_brfss$tobacco_use, 
                                  levels = reuse_yn)


short_brfss$alc_drnk_30days <- factor(short_brfss$alc_drnk_30days, 
                                      levels = reuse_yn)


short_brfss$diabetes_status <- factor(short_brfss$diabetes_status, 
                                      levels = reuse_yn)
```

```{r}
#| echo: false
#| output: false
# Check levels of each variable
lapply(short_brfss[, c("diabetes_status", "physical_activity", "smoking", 
                       "tobacco_use", "alc_drnk_30days")], table)
```

## Data cleaning: handling missing values

Checking for NAs in all columns before making the decision.

```{r}
# Check NAs in all columns at once
sapply(short_brfss, function(x) sum(is.na(x)))
```

Given that most predictor variables had large number of missing values, "Complete Case Analysis", involving removing entire rows(cases) that contain any missing values would not be appropriate as this would reduce the sample size and also it would introduce bias as these data are not missing completely at random. Therefore, we would do "Available Case Analysis" for the predictor variables where we use the data that are present for a respondent and skips for data that are missing. Example: For specific visualizations requiring selected variables, rows with missing values for all selected variables were dropped there to ensure comparative presentation. (Ref: <https://www.tandfonline.com/doi/full/10.1080/10696679.2024.2376052#d1e575>)

```{r}
#| echo: false
#| output: false
# Assigning labels to variables 
# This is placed here because with earlier variables transformation overwrote 
# the objects and removed the labels.
labs <- c(id_no = "Sequence number", 
          state_name = "Name of the State",
          age = "Age of the respondent",
          sex = "Sex of the respondent",
          income_level= "Annual Household income from all sources",
          BMI = "Body Mass Index",
          physical_activity = "Any physical activity in past month?",
          smoking = "Current smoking status",
          tobacco_use = "Current tobacco use",
          alc_drnk_30days = "At least one drink of alcohol in the past 30 days",
          diabetes_status = "Diabetes status",
          diab_first_known = "Age when first told had diabetes")

label(short_brfss) <- as.list(labs[match(names(short_brfss),
                                         names(labs))])
```

Further, the dataset was saved to .rds format to preserve the metadata of factor levels and value levels.

```{r}
#| echo: false
#| output: false
# Export cleaned dataset for further analysis, the dataset was saved to .rds 
# format to preserve the metadata of factor levels and value labels.
saveRDS(short_brfss, 
        file = "./01_Datasets/diabetes_brfss_cleaned.rds", 
        compress = FALSE)
```

## Descriptive Statistics

```{r}
#| echo: false
#| output: false
##df <- readRDS("./01_Datasets/diabetes_brfss_cleaned.rds")
df <- short_brfss
```

### 1. Socio-demographic characteristics:

#### 1.1 Population pyramid (Age & Sex): 

A population pyramid is selected because, although age was recorded as a numeric variable, I later realized that individuals above 80 years were reported simply as "\>80 years." Calculating and reporting the mean or median age would therefore not accurately represent the study population. A population pyramid is a more appropriate option, as it allows us to clearly display the age distribution, including the group above 80 years.

```{r}
# Population pyramid (Age & Sex)
# Create age groups
df$age_g <- cut(df$age,
                breaks = c(18, 25, 30, 
                           35, 40, 45, 
                           50, 55, 60, 
                           65, 70, 75, 
                           80, Inf),
                labels = c("18-24", "25-29", 
                           "30-34", "35-39", 
                           "40-44", "45-49",
                           "50-54", "55-59", 
                           "60-64", "65-69", 
                           "70-74", "75-79", 
                           ">= 80"),
                right = FALSE)


# Creating dataframe with age groups and sex for pyramid
py_df <- as.data.frame(table(df$age_g, 
                             df$sex))
colnames(py_df) <- c("age", 
                     "sex", 
                     "count")
# Count observation for n
total_pop <- sum(py_df$Freq)

# Convert count to percentage
py_df$count <- py_df$count / sum(py_df$count) * 100

# Making one sex values negative for pyramid effect
py_df$count <- ifelse(py_df$sex == "Male", 
                      -py_df$count, py_df$count)

# Create population pyramid
ggplot(py_df, 
       aes(x = age, 
           y = count, 
           fill = sex)) +
  geom_bar(stat = "identity") +
  
  # Adding percentage to bars
  geom_text(aes(label = paste0(round(abs(count), 1), "%"),
                hjust = ifelse(sex == "Male", 1.2, -0.2)),
             position = "identity",
                 size = 2.5,
                color = "black") +
  
  scale_y_continuous(labels = function(x) paste0(abs(x), "%"),
                     limits = max(py_df$count) * c(-1.2, 1.2)) +
  
  labs(title = expression(underline("Population Pyramid of BRFSS in 2023")),
    subtitle = expression(underline("n = 433,323")),
           x = "Age (in years)", 
           y = "Population (in Percentage)",
     caption = "Figure 1: Population pyramid of BRFSS in 2023") +
  
  scale_colour_manual("Sex", values = c("steelblue", 
                                        "pink"),
                    aesthetics = c("colour", 
                                   "fill")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, 
                              size = 12),
    axis.title = element_text(size = 8),
    plot.subtitle = element_text(hjust = 0.5,
                                  size = 8, 
                                 color = "gray40"),
    axis.text = element_text(size = 8),
    plot.caption = element_text(hjust = 0.5, 
                                 face = "italic", 
                                 size = 8),
    legend.position = "bottom",
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 8)) +
  coord_flip()
```
Interpretation: There were more older adults (around 50 years and above) compared to younger adults in the sample. In addition, females were more represented than males, especially in the older age groups.

### 2. Biological Characteristics:

#### 2.1 Mean BMI of the respondents

```{r}
summary(df$BMI, na.rm = TRUE)
```

Mean BMI of the respondents is 28.5, stating, in average, respondent were overweight.

### 3. Behavioral Characteristics: 

#### 3.1 2×2 panel (multi-panel) barplot 

```{r}
# Creating dataframe with behavioral variables
behav_char <- df[c("physical_activity", 
                   "smoking", 
                   "tobacco_use", 
                   "alc_drnk_30days")]

# Removing rows with NAs from all variables to make plots comparable
behav_char_clean <- behav_char[complete.cases(behav_char), ]

# Count observation for n
# nrow(behav_char_clean)

# Counts table
tab_behav <- sapply(behav_char_clean, table)

# Transpose to wide data frame
t_tab_behav <- as.data.frame(t(tab_behav))
t_tab_behav$variable <- rownames(t_tab_behav)

# Reshape to long
l_tab_behav <- reshape(t_tab_behav,
                       varying   = c("Yes", "No"),
                       v.names   = "Count",
                       timevar   = "Response",
                       times     = c("Yes", "No"),
                       direction = "long")
# Set factor orders
l_tab_behav$Response <- factor(l_tab_behav$Response, levels = c("Yes", "No"))
l_tab_behav$variable <- factor(l_tab_behav$variable,
                               levels = c("physical_activity", "smoking",
                                          "tobacco_use", "alc_drnk_30days"),
                               labels = c("Physical activity in month", "Smoking",
                                          "Tobacco use", "Alcohol intake in 30 days"))
# Bar plot
ggplot(l_tab_behav, aes(x = Response, y = Count, fill = Response)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Yes" = "#1f77b4", "No" = "orange")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.12))) +
  facet_wrap(~ variable, nrow = 2, ncol = 2) +
  geom_text(aes(label = scales::comma(Count)),
            vjust = -0.4, size = 3.2) +
  labs(title = expression(underline("Behavioral Characteristics: BRFSS in 2023")),
    subtitle = expression(underline("n = 399,230")),
           x = NULL, 
           y = "Count",
     caption = "Figure 2: Behavioral characteristics of the respondents of BRFSS in 2023")  +
  theme_minimal(base_size = 12) +
  theme(
        plot.title = element_text(hjust = 0.5, 
                                   size = 12),
        plot.subtitle = element_text(hjust = 0.5,
                                  size = 8, 
                                 color = "gray40"),
        axis.title = element_text(size = 8),
        axis.text = element_text(size = 8, 
                                hjust = 1),
        plot.caption = element_text(hjust = 0.5, 
                                     face = "italic", 
                                     size = 8),
        legend.position    = "none",
        strip.text         = element_text(face = "bold"),
        panel.grid.major.x = element_blank())

```
Interpretation: Most respondents reported being physically active at least once in the past month. When it comes to smoking and tobacco use, more people identified as smokers compared to those who used tobacco products directly. Additionally, the majority of respondents reported consuming alcohol in the past 30 days. 

### 4. Diabetes related: 

#### 4.1 Diabetes status by income level

```{r}
# Keep filled rows in both income_level and diabetes_status, removing NAs
income_diab <- df[c("income_level", "diabetes_status")]
income_diab_clean <- income_diab[complete.cases(income_diab), ]
# Count observation for n
# nrow(income_diab_clean)

#plot
ggplot(income_diab_clean, aes(x = income_level, fill = diabetes_status)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Yes" = "darkred", "No" = "#4daf4a")) +
  labs(
    title = expression(underline("Diabetes Status by Income Level: BRFSS, 2023")),
    subtitle = expression(underline("n = 346,224")),
    caption = "Figure 3: Diabetes status by income level of respondents of BRFSS in 2023",
    x = "Income Level",
    y = "Frequency",
    fill = "Diabetes") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, 
                              size = 12),
    axis.title = element_text(size = 8),
    plot.subtitle = element_text(hjust = 0.5,
                                  size = 8, 
                                 color = "gray40"),
    plot.caption = element_text(hjust = 0.5, 
                                 face = "italic", 
                                 size = 8),
    axis.line = element_line(color = "grey", 
                          linewidth = 0.8),
    axis.text.x = element_text(size = 8, 
                            angle = 45,
                            hjust = 1),
    axis.text.y = element_text(size = 8))
```
Interpretation: Across all income levels, most respondents did not have diabetes, with the majority of the sample falling in the middle to higher income range. However, lower income groups carried a disproportionately higher burden of diabetes relative to their size.

#### 4.2 US States by Diabetes Prevalence: BRFSS, 2023

```{r}
# Visualize the diabetes prevalence on US MAP
# Create dataframe with state name and diabetes cases
diab_st_df <- as.data.frame(table(df$state_name, df$diabetes_status))
colnames(diab_st_df) <- c("state", 
                     "diabetes_status", 
                     "frequency")

# Widening the df with separate "No" count
diab_st_df <- reshape(diab_st_df, 
                     idvar = names(diab_st_df)[1],
                     timevar = "diabetes_status", 
                     direction = "wide",
                     v.names = "frequency")

# Stripping the "frequency." prefix
names(diab_st_df) <- gsub("frequency\\.", "", names(diab_st_df))

# Adding total column
diab_st_df$total <- diab_st_df$Yes + diab_st_df$No

# Adding prevalence column
diab_st_df$prevalence <- (diab_st_df$Yes / diab_st_df$total) * 100

# Before plotting, verify all U.S. states are represented in diab_state_df
# by comparing against the built-in 'statepop' reference dataset from the
# usmap package.

# statepop <- usmap::statepop
# setdiff(statepop$full, diab_st_df$state)
# There is no diabetes data available from the "Kentucky", "Pennsylvania" state.
# Sometimes the data are not reported for certain states if they don't fulfill 
# certain requirements.

# Prep to add each state name to label on map

# Plotting the diabetes cases from survey in the US map
plot_usmap(data = diab_st_df,
           regions = "states",
           values = "prevalence",
           include = c(state.abb, "PR"), #default: excludes Puerto Rico
           color = "white",
           labels = TRUE) +  
  scale_fill_continuous(low = "lightyellow", high = "darkred") +
  labs(
    title = expression(underline("US States by Diabetes Prevalence: BRFSS, 2023")),
    subtitle = expression(underline("n = 433,323")),
    fill = "Prevalence",
    caption = "Figure 4: US states by Diabetes Prevalence: BRFSS, 2023") +
  theme(
    plot.title = element_text(hjust = 0.5, 
                               size = 15, 
                               face = "bold",
                              margin = margin (b = 5)),
    plot.subtitle = element_text(hjust = 0.5,
                                  size = 8, 
                                 color = "black"),
    plot.caption = element_text(hjust = 0.5, 
                                 face = "italic", 
                                 size = 8),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 40),
    plot.background = element_rect(fill = "lightblue", 
                                  color = NA),
    legend.position = "right",
    legend.background = element_rect(fill = "lightblue", 
                                    color = NA))
```
Interpretation: Diabetes prevalence varied across US states, with the southeastern region showing the highest rates, particularly in states like West Virginia, and Alabama. In contrast, western and northern states such as Colorado, Utah, and Montana had notably lower diabetes prevalence. There is no diabetes data available from the "Kentucky", "Pennsylvania" state.

#### 4.3 Mean Age of First Known Diabetes Diagnosis

```{r}
mean_diab_first <- mean(df$diab_first_known, na.rm = TRUE)
```

On average, individuals in this dataset first became aware of their diabetes diagnosis at approximately 53.40 years of age.

## Multivariable Logistic Regression Analysis

Since we are interested to test the association of behavioral variables with the Diabetes, only relevant variables were administered to the model. Further, logistic regression predicts the probability of the higher level, hence, for better interpretation of the results, the level were reordered as "no" and "yes".

```{r}
#table(df$diabetes_status)
df$diabetes_status <- factor(df$diabetes_status, levels = c("No", "Yes"))

# Logistics regression calculation
LR <- glm(diabetes_status ~ physical_activity + smoking + tobacco_use +
              alc_drnk_30days, 
            family = binomial(link = "logit"), data = df)

# Report estimated odds ratio,p-value and 95% CI. 
jtools::summ(LR, exp = T, confint = T, model.fit = F, digits = 3)
```

Interpretation: According to multivariable logistic regression, the odds of having diabetes among those who are not physically active in a month increases by 2.035 times (95% CI:1.996 to 2.074; p-value \< 0.001). Similarly, the odds of having diabetes among those who drink alcohol in 30 days increases by 2.015 times (95% CI: 1.978 to 2.053; p-value \< 0.001). Interestingly, model showed that current non-smokers had statistically significant likelihood of having diabetes by 1.097 times (95% CI: 1.066 to 1.129; p-value \< 0.001) compared to those who did not smoke. However, there is no statistical significant association of current tobacco use with having diabetes OR = 0.981 (95% CI: 0.929 to 1.037; p-value = 0.502). 


## Conclusion
The report explored socio-demographic, biological, behavioral and diabetes related characteristics. Among the behavioral characteristics, lack of physical activity in a month and recent alcohol consumption, had statistical significant association with the diabetes. Additionally, current non-smoking status was also associated with having diabetes which could reflect behavioral modification after diagnosis. 
